Guidelines for reliable and reproducible functional enrichment analysis

### 2021-12-07 analysis_code/QC_analysis2.Rmd
tidy3

dir.create("images")
png("images/journals2.png",width=400,height=300)
par(mar=c(5,12,3,3))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Journal", xlim=c(0,170))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 10, label = res2, pos = 3, cex = 1, col = "black")
dev.off()
```

@@

@@

@@

@@

@@

@@

png("images/fdr2.png",width=300,height=200)
par(mar=c(5,8,3,3))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "FDR correction performed", xlim=c(0,160))
text(y = (1:length(res)*1.2)-1.2  , x = res + 15, label = res, pos = 3, cex = 1, col = "black")
dev.off()
```

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

dir.create("images")
png("images/journals2.png",width=400,height=300)
par(mar=c(5,12,3,3))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Journal", xlim=c(0,170))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 10, label = res2, pos = 3, cex = 1, col = "black")
dev.off()

pdf("images/journals2.pdf",width=4,height=4)
par(mar=c(5,9,3,2))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Journal", xlim=c(0,180))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 15, label = res2, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/omics2.pdf",width=4,height=4)
par(mar=c(5,10,3,2))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Omics type", xlim=c(0,110))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 10, label = res2, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/organisms2.pdf",width=4,height=4)
par(mar=c(5,11,3,2))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Organism", xlim=c(0,200))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 25, label = res2, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/genesetlib2.pdf",width=4,height=4)
par(mar=c(5,11,3,2))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene set library", xlim=c(0,150))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 16, label = res2, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/genesetvers2.pdf",width=3,height=2)
par(mar=c(5,4,3,2))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene set version defined", xlim=c(0,260))
text(y = (1:length(res)*1.2)-1.2  , x = res + 25, label = res, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/stattest2.pdf",width=4,height=4)
par(mar=c(5,11,3,2))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Test used", xlim=c(0,170))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 20, label = res2, pos = 3, cex = 1, col = "black")
dev.off()

@@

png("images/fdr2.png",width=300,height=200)
par(mar=c(5,8,3,3))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "FDR correction performed", xlim=c(0,160))
text(y = (1:length(res)*1.2)-1.2  , x = res + 15, label = res, pos = 3, cex = 1, col = "black")
dev.off()

pdf("images/fdr2.pdf",width=3,height=3)
par(mar=c(5,5,3,2))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "FDR correction performed", xlim=c(0,160))
text(y = (1:length(res)*1.2)-1.0  , x = res + 15, label = res, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/app2.pdf",width=4,height=4)
par(mar=c(5,12,3,2))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "App used", xlim=c(0,80))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 10, label = res2, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/appvers2.pdf",width=3,height=2)
par(mar=c(5,4,3,2))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "App version defined", xlim=c(0,200))
text(y = (1:length(res)*1.2)-1.2  , x = res + 19, label = res, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/code2.pdf",width=3,height=2)
par(mar=c(5,4,3,2))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Code availability", xlim=c(0,55))
text(y = (1:length(res)*1.2)-1.2  , x = res + 5, label = res, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/bg2.pdf",width=4,height=3)
par(mar=c(5,10,3,2))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Background list specified", xlim=c(0,220))
text(y = (1:length(res)*1.2)-1.1  , x = res + 20, label = res, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/genelists2.pdf",width=3,height=2)
par(mar=c(5,4,3,2))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene lists provided", xlim=c(0,170))
text(y = (1:length(res)*1.2)-1.25  , x = res + 15, label = res, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/assumptions2.pdf",width=4,height=3)
par(mar=c(5,11,3,2))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Methodological flaws", xlim=c(0,230))
text(y = (1:length(res)*1.2)-1.2  , x = res + 25, label = res, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/comb2.pdf",width=3,height=2.5)
par(mar=c(5,5,3,2))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Separated or combined ORA", xlim=c(0,100))
text(y = (1:length(res)*1.2)-1.1  , x = res + 10, label = res, pos = 3, cex = 1, col = "black")
dev.off()

@@

pdf("images/gseaparam2.pdf",width=4.5,height=3)
par(mar=c(5,6,3,2))
barplot(t(params),horiz=TRUE,las=1,xlim=c(0,30),legend = colnames(params),
        main="GSEA parameter reporting", xlab="no. analyses",
        args.legend = list(x = "topleft", bg="white",inset = c(0.05, 0.1)))
text(y = (1:nrow(params)*1.2) - 0.8 , x = params[,1]-3, label = params[,1], pos = 3, cex = 1, col = "white")
text(y = (1:nrow(params)*1.2) - 0.8 , x = params[,1]+2, label = params[,2], pos = 3, cex = 1, col = "black")
dev.off()

### 2021-12-07 analysis_code/analysis.Rmd
tidy3

if (file.exists("cites_data.rds") ) {
  cites <- readRDS("cites_data.rds")
} else {
  PMCIDS <- as.numeric(gsub("PMC","",x$Pubmed.Central.ID ))
  mylink <- elink(uid=PMCIDS , dbFrom="pmc", dbTo = "pubmed")
  PMID <- mylink[1]
  #writeLines(PMID,"pubmedids.txt")
  res <- sapply(X=PMID,FUN=function(x){
      Sys.sleep(1)
      CITES <- elink(uid = x , dbFrom = "pubmed"  , linkname = "pubmed_pubmed_citedin" ) 
      CITES=length(CITES[1])
      return(CITES)
   })
  mylink2 <- elink(uid=PMID , dbFrom="pubmed", dbTo = "pmc")
  cites <- data.frame(mylink2[1],PMID,res)
  colnames(cites) <- c("PMCID","PMID","cites")
  saveRDS(cites,file="cites_data.rds")
}

cites$PMCID <- gsub("^","PMC",cites$PMCID)
cites$PMID <- gsub("^","PMID",cites$PMID)
xm <- merge(x,cites,by.x="Pubmed.Central.ID",by.y="PMCID")
dim(xm)

@@

mylm2 <- lm (xm$scores ~ xm$cites)
plot(xm$cites,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
  xlab="Num cites",ylab="score",main="Association of analysis scores with number of citations")
abline(mylm2,col="red")

cor.test(xm$scores , xm$cites, method="pearson")

cor.test(xm$scores , xm$cites, method="spearman")

mylm2 <- lm (xm$scores ~ log2(xm$cites))
plot(log2(xm$cites),xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
  xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")
abline(mylm2,col="red")

png("images/cites1.png",width=350,height=300)
par(mar=c(5,5,3,1))
plot(log2(xm$cites),xm$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
  xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")
abline(mylm2,col="red")
dev.off()

cor.test(xm$scores , xm$cites, method="pearson")

cor.test(xm$scores , xm$cites, method="spearman")

pdf("images/cites1.pdf",width=3.5,height=3)
par(mar=c(5,5,3,1))
plot(log2(xm$cites),xm$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
  xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")
abline(mylm2,col="red")
dev.off()

>>>>

if (file.exists("cites_data.rds") ) {
  cites <- readRDS("cites_data.rds")
} 
cites$PMID <- gsub("^","PMID",cites$PMID)
xm <- merge(x,cites,by.x="Pubmed.Central.ID",by.y="PMCID")
dim(xm)
head(xm)

@@

mylm2 <- lm (xm$scores ~ xm$NumCites)
plot(xm$NumCites,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
  xlab="Num cites",ylab="score",main="Association of analysis scores with number of citations")
abline(mylm2,col="red")

cor.test(xm$scores , xm$NumCites, method="pearson")

cor.test(xm$scores , xm$NumCites, method="spearman")

mylm2 <- lm (xm$scores ~ log2(xm$NumCites+1))
plot(log2(xm$NumCites),xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
  xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")
abline(mylm2,col="red")

png("images/cites1.png",width=350,height=300)
par(mar=c(5,5,3,1))
plot(log2(xm$NumCites),xm$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
  xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")
abline(mylm2,col="red")
dev.off()

cor.test(xm$scores , xm$NumCites, method="pearson")

cor.test(xm$scores , xm$NumCites, method="spearman")

pdf("images/cites1.pdf",width=3.5,height=3)
par(mar=c(5,5,3,1))
plot(log2(xm$NumCites),xm$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
  xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")
abline(mylm2,col="red")
dev.off()

### 2021-12-07 analysis_code/citation_fetching.Rmd
tidy3

get_cites <- function(PMID) {
  CITES <- cites(PMID)
  N_CITES <- length(CITES[1])
  return(c(PMID,N_CITES))
}

@@

@@

Except reutils fails for articles with any special characters in there.

@@

saveRDS(df,"cites_data.rds")

>>>>

get_cites <- function(PMID) {
  CITES <- cites(PMID)
  N_CITES <- length(linkset(CITES)[1]$pubmed_pubmed_citedin)
  return(c(PMID,N_CITES))
}

@@

# test a paper with 0 cites
PMID=31148949
res <- get_cites(PMID = PMID)
res

@@

Except reutils fails for articles with any special characters in there, looks like it is causedby improper handling of non
ASCII characters by the XML parser.
Solution is to use JSON format.

@@

plot(df$Pubdate,df$NumCites)
mylm <- lm(df$NumCites ~ df$Pubdate)
abline(mylm,col="red")
summary(mylm)
cor.test(df$NumCites , as.numeric(df$Pubdate))
saveRDS(df,"cites_data.rds")

### 2021-12-07 docs/manuscript.Rmd
tidy3

* Gene set library used, and whether a version was reported

@@

* Software package used, and whether a version was reported

@@

* Whether gene profile was provided in the supplement

>>>>

* Gene set library used, and whether a version or date was reported

@@

* Software package used, and whether a version or date was reported

@@

* Whether gene profile was provided (in the supplement or main text).
